Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-25 submission/manuscript.Rmd
finish discussion draft

- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na&#239;ve to primed pluripotency.
	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.

		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
	- However, Recently KDM5C's catalytic acivity was found to be unnecessary for dazl suppression in ESCs.  
	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes. 
	- This would be the first (?) example in which removal of an active mark is required for germline gene repression.

- In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain. 
	- In EpiLCs, KDM5C directly represses key drivers of germline identity like Dazl, likely through PRC1.6 recruitment and promoting CpG methylation. However, _Kdm5c_-KO also ectopically expresses germline genes activated indirectly, including _Stra8_.  
	- The germline developmental program to some extent continues ectopically during _Kdm5c_-KO development, resulting in aberrant transcription of late stage spermatogenesis genes later in life. 
	- These results define KDM5C's role in the demarcation between soma and gemrline identity and offer a window into potential targets to assess the deletirous ecfects these ectopic genes on neurodevelopment.

>>>>

It is hypothesized distinct repressive chromatin marks are placed at germline gene promoters as the embryo transitions from na&#239;ve to primed pluripotency. Initially, germline genes are repressed by placement of hisone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3) in ESCs and then gain _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. KDM5C may be instead required to remove an active mark, histone 3 lysine 4 trimethylation (H3K4me3), since H3K4me3 can impede _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found an increase in H3K4me3 around the transcription start site (TSS) of germline DEGs and loss of CpGme at at least two germline promoters[@scandagliaLossKdm5cCauses2017]. However, KDM5C's role in embryonic germline gene silencing is still unclear, ask KDM5C's catalytic activity was recently found to be dispensible for silencing DAZL in ESCs[@guptaGenomewideScreenReveals2023]. In this study, we observed a global increase in H3K4me3 around the TSS of germline-enriched gene in the _Kdm5c_-KO amygdala and an increase in H3K4me2 in _Kdm5c_-KO EpiLCs. We found KDM5C's expression is dynamically regulated during ESC to EpiLC differentiation and loss of KDM5C leads to impaired placement of CpGme in extended EpiLCs. Altogether, this suggests KDM5C is necessary during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.

In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain. In EpiLCs, KDM5C directly represses key drivers of germline identity like _Dazl_, by promoting intial CpG methylation placement in the post-implantation embryo. However, germline genes can also become ectopically expressesed in _Kdm5c_-KO cells independent of direct KDM5C regulation, including the meiotic transcription factor _Stra8_. These ectopic germline developmental programs can, to some extent, mimic typical germ cell development, resulting in aberrant transcription early developmental and meiotic genes in _Kdm5c_-KO EpiLCs and late-stage spermatogenesis genes in the _Kdm5c_-KO brain. Altogether, these results define KDM5C's role in the demarcation between soma and gemrline identity and offers novel insight into how this dysregulation of tissue identity changes over the course of development. Additionally, this study provides the mechanistic foundation required to ultimately investigate the impact of aberrant germline identity upon neurodevelopment.

- include cancer in there somewhere (Somatic misexpression of germline genes has been implicated in many cancers.)
